\begin{abstract}
Reconstructing the three-dimensional (3D) distribution of the intergalactic medium (IGM) from sparse one-dimensional (1D) Lyman-alpha forest sightlines is a sparse-view tomography problem with a physically derived forward operator. Classical voxel-grid methods such as TARDIS~\cite{horowitz2019tardis} and Wiener filtering~\cite{stark2015tomography} discretize the volume up front, trading resolution for memory or memory for spectral aliasing.

We present \emph{CosmoGasVision}, a continuous neural field $F_\theta:\mathbf{x} \to (\rho, T, X_{HI}, v_{\text{pec}})$ optimized end-to-end against a fully differentiable Lyman-alpha optical-depth integrator. The integrator implements the Tepper-Garc\'ia~\cite{tepper2006voigt} analytic Voigt-Hjerting kernel, full redshift-space-distortion (RSD) convolution, a learnable amplitude that absorbs absolute calibration constants, a saturated-absorber mask~\cite{wolfe2005}, and a mean-flux soft anchor at the verified value $\langle F\rangle_{\text{obs}}(z=0.3) = 0.979 \pm 0.005$~\cite{kirkman2007}.

At the corrected anchor and full publication schedule (\texttt{pub-t1}, $50{,}000$ steps, four Sherwood~\cite{bolton2017sherwood} feedback variants $P_1$--$P_4$, seed $=0$), \emph{two of three} primary [D-13] gates close at the single-seed level: mean-flux closes $4/4$ at the seed=42 single-seed anchor (cross-physics spread $0.63\%$, $<\!1\%$ bar), but a 5-seed sightline-level bootstrap shows the population CI overlaps the Kirkman+2007 anchor $[0.974, 0.984]$ in only $2/4$ cells ($P_3$ and $P_4$ marginal at q84); $P_1$ and $P_2$ bootstrap CIs sit entirely below the anchor band. The flux-PDF Kolmogorov--Smirnov gate closes $3/4$ bootstrap-robustly. The binding gate fails: the 1D flux-power-spectrum residual $|\Delta P_F/P_F|$ over the inertial range $k_\parallel \in [10^{-2.5}, 10^{-1.5}]$ s/km remains $36\%$--$42\%$ in $4/4$ cells, $3.6$--$4.2\times$ over the $10\%$ bar. A $\tau$-binned residual decomposition cross-physics positively identifies saturation-regime under-fitting ($F \in [0.05, 0.30]$) as the residually-supported mechanism (concentration ratio $R \in [18.87, 23.75]$, $>\!12\times$ over the positive-ID floor in $4/4$ cells). Three counterfactual interventions against this mechanism --- a saturation-aware $P_F$ band loss, an FGPA-tail per-pixel regularizer, and velocity-gradient conditioning --- each failed by a previously-unaudited degeneracy class, retired under pre-committed stop conditions.

The contributions are (i) a continuous-neural-field formulation of Ly$\alpha$ tomography coupled to a differentiable Voigt integrator, with a cost-survey-validated linear-VRAM training schedule across feedback physics; (ii) the calibration-vs-structure falsification that surfaces $P_F$ --- not anchor calibration, not sightline density alone --- as the binding gate, with the deficit residually localized to the saturation band cross-physics; (iii) a falsification-disciplined methodology of pre-committed gates and degeneracy-aware counterfactuals that converts negative results into successor-experiment constraints. We do not claim to have solved 3D IGM reconstruction; we report a partial result with the binding obstacle named.
\end{abstract}
