\section{Introduction}
\label{sec:intro}

This paper is, in computer-vision language, a sparse-view tomography problem with an unusual sensing modality. The volume to be reconstructed is the \textbf{intergalactic medium} (IGM) --- the diffuse gas filling the space between galaxies, which holds the majority of the universe's ordinary (baryonic) matter and forms the filaments and voids of the \emph{cosmic web}. We have no camera that can image this volume directly. Instead, we observe it through \textbf{Lyman-alpha forest} spectra: when light from a distant quasar passes through the IGM on its way to Earth, neutral hydrogen along the path absorbs photons at a specific resonance wavelength, and Doppler shifts caused by cosmic expansion smear that absorption into a dense forest of narrow dips across the quasar's spectrum. Each such 1D spectrum is a \textbf{sightline}: a single ray through the cosmological volume --- the analogue of one CT scanline through a body, or one pixel column in a 3D scan --- and the position of every absorption dip encodes the gas state at one location along that ray.

Reconstructing the 3D IGM from a sparse bundle of such 1D sightlines is therefore literally tomography from pencil-beam projections. In astrophysics it is referred to as \emph{Lyman-alpha tomography}~\cite{stark2015tomography}; in CV vocabulary it is sparse-view inverse rendering with a known, physically-derived forward operator. The astrophysical motivation is strong: a recovered 3D map at the resolution of upcoming surveys (e.g., DESI) is the substrate for studying structure formation, reionization, and how galaxies feed energy back into their surroundings.

Classical solutions like TARDIS~\cite{horowitz2019tardis} and Wiener filtering~\cite{stark2015tomography} discretize the volume into a fixed voxel grid. This is the same trade-off familiar from CT and from grid-based NeRF predecessors: at the typical inter-sightline spacing of survey data (several $h^{-1}$ Mpc transverse, where $h^{-1}$ Mpc is a comoving-distance unit that factors out cosmic expansion), a voxel grid either over-commits memory or aliases the small-scale absorption structure that the metrics actually care about.

\textbf{Motivation.} Neural Radiance Fields (NeRF)~\cite{mildenhall2020nerf} replaced explicit voxel grids in vision with a coordinate-conditioned MLP, supervised through a differentiable rendering integral. The same representational move is what the IGM problem needs. We are not interested in rendering visual color; we are interested in recovering an underlying physical state from absorption profiles, and we want a representation whose gradient flows cleanly through the forward physics rather than getting interrupted by a discretization seam.

\textbf{Contribution.} We introduce \emph{CosmoGasVision}, a NeRF-style continuous field for cosmological gas, coupled to a custom differentiable projector that maps the MLP-decoded physical fields to 1D Lyman-alpha optical depth (a logarithmic measure of light absorbed along the sightline; see \cref{sec:method}) using the analytic Voigt-Hjerting kernel~\cite{tepper2006voigt}. The contribution is the representation-plus-forward-model package: a continuous IGM field optimized end-to-end against the same physics that astrophysicists already trust, so that recovered structures are sub-voxel in resolution while every step of the gradient path remains physically interpretable.

\textbf{What this paper is, and is not.} The CVPR community is owed an unambiguous statement of claim ceiling. This paper is \emph{not} a closed-loop demonstration that a continuous neural field solves 3D IGM tomography from sparse Ly$\alpha$ sightlines. At the corrected mean-flux anchor and the full publication schedule across four Sherwood feedback variants ($P_1$--$P_4$), two of three primary [D-13] structure gates close at the single-seed level --- mean-flux ($4/4$ at the seed=42 single-seed anchor, cross-physics spread $\meanFseedFortyTwoSpread$ at a $<\!1\%$ bar; under a 5-seed sightline-level bootstrap the population CI overlaps the Kirkman+2007 anchor $[\AnchorMeanFLo, \AnchorMeanFHi]$ in only $2/4$ cells, $P_3$/$P_4$ marginal at q84, [D-44]) and the flux-PDF Kolmogorov-Smirnov gate ($3/4$ bootstrap-robust, the breach at $P_2$ at $1.5\times$ the bar) --- while the third, the inertial-range residual $|\Delta P_F/P_F|$ of the 1D flux power spectrum, fails in $4/4$ cells at $\PFfactorOverGateLo\times$--$\PFfactorOverGateHi\times$ the gate. A $\tau$-binned residual decomposition (Sec.~\ref{sec:experiments:tau-binned}) positively identifies the saturation regime $F \in [\SatBandLo, \SatBandHi]$ cross-physics as the residually-supported mechanism (concentration ratio $R \in [\RsatRangeLo, \RsatRangeHi]$, $4/4$ cells $>\!\RsatFactorOverFloor\times$ the positive-ID floor). Three counterfactual interventions against that mechanism --- a saturation-aware $P_F$ band loss~([D-40]), an FGPA-tail per-pixel regularizer~([D-41]), and a velocity-gradient conditioning branch~([D-42]) --- each retired by a previously-unaudited degeneracy: integrated-statistic constant-prediction shrink, self-consistent state collapse, and partial-collapse signature, respectively. The remaining gap is therefore residually supported but not closed; the strict ``saturation under-fitting \emph{is the cause} of the $P_F$ gap'' claim is reserved for successor work.

\textbf{What this paper is.} (i) A continuous-neural-field formulation of Ly$\alpha$ tomography with a differentiable RSD-convolved Voigt integrator, a saturated-absorber mask, and a mean-flux soft anchor at the verified low-$z$ value~\cite{kirkman2007}; (ii) a cost-survey-validated training schedule with a closed-form linear-VRAM model across feedback physics and sightline-density tiers; (iii) a calibration-vs-structure falsification that surfaces $P_F$ --- not anchor calibration, not sightline density alone --- as the binding gate, with the deficit residually localized to the saturation band cross-physics; (iv) a falsification-disciplined methodology of pre-committed structure gates, degeneracy-aware counterfactuals, and pre-committed stop conditions on each counterfactual, so negative results become explicit constraints on the successor design rather than discarded attempts. We compare to the classical voxel-grid baselines TARDIS~\cite{horowitz2019tardis} and Wiener filtering~\cite{stark2015tomography} qualitatively rather than head-to-head; a quantitative bake-off at matched evaluation gates is sequenced behind a $P_F$-passing reconstruction.
