\section{Experimental Setup}
\label{sec:experiments}

\subsection{Dataset}
We use the \textbf{Sherwood Simulation Suite}~\cite{bolton2017sherwood} 60 Mpc/h Planck-cosmology volume at $2048^3$ resolution; the dev redshift is $z = 0.3$. Each snapshot ships 16{,}384 sightlines along a single axis at 2048 velocity bins per ray, with ground-truth $\tau_{HI}(v)$ stored as \texttt{tauH1\_2048\_n16384\_z0.300.dat}. Sherwood ships four feedback variants: $P_1$ (no feedback), $P_2$ (stellar wind), $P_3$ (wind $+$ AGN), $P_4$ (wind $+$ strong AGN), an ablation matrix over baryonic feedback prescriptions. We train one independent field per variant so a downstream feedback-classification probe cannot leak the label through a conditioning input. We report Tiers 1--3 across all four variants; Tier 4 is deferred per the compute argument of \S\ref{sec:experiments:compute}.

\subsection{Evaluation Metrics}
Three primary structure gates are computed against the simulation ground truth on the rendered $F=e^{-\tau}$:
\begin{itemize}
    \item \textbf{1D flux power spectrum} $P_F(k_\parallel)$ (Hann-windowed periodogram, $dv/\sum w^2$ leakage compensation): $|\Delta P_F / P_F| < \GatePF$ averaged over $k_\parallel \in [10^{\PFkLoExp}, 10^{\PFkHiExp}]$ s/km~\cite{walther2018powerspectrum, boera2019thermal}.
    \item \textbf{Density-density cross-correlation} $\xi_{\hat\rho, \rho}(r)$ (Pearson on $\delta=\rho/\bar\rho-1$, FFT cross-power, periodic 60 Mpc/h grid, spherical-shell binning): $\xi_{\hat\rho, \rho}(r{=}2\,h^{-1}\,\text{Mpc}) > \GateXi$ at fiducial~\cite{lee2015, stark2015tomography}.
    \item \textbf{Flux PDF}: two-sample KS distance on $F$ restricted to $F \in [\PDFwinLo, \PDFwinHi]$ (lower cut excludes saturated absorbers; upper cut excludes continuum/metal-line residuals near $F\approx 1$~\cite{lee2015}): KS $< \GateKS$.
\end{itemize}
The fiducial pass point is $P_1$, $z = 0.3$, $n_{\text{rays}} = 1024$. The differentiable integrator was re-validated end-to-end at production architecture on the AWS SageMaker \texttt{ml.g5.xlarge} loop with S3 checkpoint sync; peak VRAM at the smallest tier is $\TonePeakVRAM$ GB, the anchor for the per-tier microbatch table of \S\ref{sec:method}.

\subsection{Stage 2b cost-survey: mean-flux anchor retraction}
\label{sec:experiments:cost-survey-sweep}
The Stage 2b cost-survey production sweep ran the per-tier schedule of \S\ref{sec:method} across all four feedback variants at Tiers 1--3 (12 banked cells, all PASS pre-registered safety rails; \texttt{peak\_vram\_gb} identical to four decimals across physics within each tier, empirically confirming a linear-VRAM scaling model). Micro-grid calibration and per-tier numerics are in the supplementary. The Stage 2b runs were trained against $\langle F\rangle_{\text{obs}} = 0.877$ at $z=0.3$. Phase~A.1 literature audit found this value corresponds to a $z\!\sim\!2$ mean transmitted flux mislabeled as $z=0.3$ during project history; the verified low-redshift value is $\langle F\rangle_{\text{obs}}(z=0.3) = \AnchorMeanF \pm \AnchorMeanFsigma$~\cite{kirkman2007}, retracted as anchor target. The structure gates ($P_F$ inertial-range residual, KS-PDF distance, $\xi_{\hat\rho,\rho}$) are anchor-invariant by construction (invariant to uniform multiplicative rescaling of $\hat F$), so the cost-survey calibration evidence is unaffected. A separate corrected-anchor full-schedule run (\texttt{pub-t1}, $50{,}000$ steps, four physics, seed $=0$) trained against $\langle F\rangle_{\text{obs}}=\AnchorMeanF$; the structure-gate readout is in \S\ref{sec:experiments:cosmological-eval}. The $\tau_{\max}$ sensitivity gate cleared at $\sim\!0.01$--$0.02\%$ (suppl.); $\tau_{\max}=10$ is locked.

\subsection{Main Results: 4-physics $\times$ 3-gate readout}
\label{sec:experiments:headline}
Tab.~\ref{tab:headline-3gate} consolidates the structure-gate readout across the four Sherwood feedback variants at the corrected-anchor full publication schedule (\texttt{pub-t1}). The reader who only reads this table should leave with the complete claim ceiling. Mean-flux closes $4/4$ at the seed=42 single-seed anchor, but the 5-seed sightline-level bootstrap CI overlaps the Kirkman+2007 anchor $[\AnchorMeanFLo, \AnchorMeanFHi]$ in only $2/4$ cells ($P_3$, $P_4$ marginal at q84). KS closes $3/4$ (bootstrap-robust; $P_2$ q84 $=\KSBootstrapQHiPtwo$ fails). $P_F$ fails $4/4$ at $\PFfactorOverGateLo$--$\PFfactorOverGateHi\times$ the gate. The 1D-along-ray density-correlation proxy is measured on the cost-survey checkpoint at $P_1$ only and is anchor-invariant by construction (Tab.~\ref{tab:headline-3gate} footnote). Training is single-seed (seed $=0$). KS and $\langle F_{\rm pred}\rangle$ are reported with 5-seed sightline-level bootstrap CIs ($K=\BootstrapK$ resamples per cell, prediction and truth-side seeds from independent PCG64 streams; anti-degeneracy F1/F2 audit PASS, suppl.). $P_F$ remains single-seed (seed $=42$ anchor, bootstrap deferred).

\textbf{Bootstrap-vs-seed=42 mean-flux divergence (symmetric-honesty disclosure).} The bootstrap mean for $\langle F_{\rm pred}\rangle$ lies systematically below the seed=42 anchor by $\meanFseedDeviationRangeLo$--$\meanFseedDeviationRangeHi\sigma$ across all four cells ($P_1$: $\meanFseedDeviationPone$, $P_2$: $\meanFseedDeviationPtwo$, $P_3$: $\meanFseedDeviationPthree$, $P_4$: $\meanFseedDeviationPfour$), indicating the single-seed reporting samples the high-$\langle F\rangle$ tail of the sightline distribution. KS values are not similarly biased. This is a property of the seed=42 sample, not a bootstrap pathology; it is disclosed here because the bootstrap softens rather than strengthens the single-seed reading, and our pre-registered honest-reporting protocol requires both directions to be reported symmetrically.

\begin{table*}[t]
    \centering
    \renewcommand{\arraystretch}{1.2}
    \setlength{\tabcolsep}{4pt}
    \footnotesize
    \begin{tabular}{l|cc|cc|cc|c}
    Cell & $\langle F_{\text{pred}}\rangle$ (mean $\pm \sigma$) & MF call & KS ($<\!\GateKS$, mean $\pm \sigma$) & KS call & $|\Delta P_F/P_F|$ ($<\!\GatePFdec$) & $P_F$ call & $r_\rho^{\log}$ ($>\!\GateXi$)$^\dagger$ \\
    \hline
    $P_1$ (no fdbk)     & $\meanFBootstrapMeanPone \pm \meanFBootstrapStdPone$ & \textbf{FAIL} (CI $<$ anchor) & $\KSBootstrapMeanPone \pm \KSBootstrapStdPone$ & \textbf{PASS} & $\mathbf{\PFresidPone}$ & \textbf{FAIL $\PFfactorOverGateHi\times$} & $\rrhoLogProxyPone$ (FAIL) \\
    $P_2$ (wind)        & $\meanFBootstrapMeanPtwo \pm \meanFBootstrapStdPtwo$ & \textbf{FAIL} (CI $<$ anchor) & $\KSBootstrapMeanPtwo \pm \KSBootstrapStdPtwo$ & \textbf{FAIL $1.5\times$} & $\mathbf{\PFresidPtwo}$ & \textbf{FAIL $3.8\times$} & --- (owed) \\
    $P_3$ (wind+AGN)    & $\meanFBootstrapMeanPthree \pm \meanFBootstrapStdPthree$ & \textbf{PASS} (q84 marginal) & $\KSBootstrapMeanPthree \pm \KSBootstrapStdPthree$ & \textbf{PASS} & $\mathbf{\PFresidPthree}$ & \textbf{FAIL $\PFfactorOverGateLo\times$} & --- (owed) \\
    $P_4$ (strong AGN)  & $\meanFBootstrapMeanPfour \pm \meanFBootstrapStdPfour$ & \textbf{PASS} (q84 marginal) & $\KSBootstrapMeanPfour \pm \KSBootstrapStdPfour$ & \textbf{PASS} & $\mathbf{\PFresidPfour}$ & \textbf{FAIL $\PFfactorOverGateLo\times$} & --- (owed) \\
    \hline
    Tally (bootstrap)   & \multicolumn{2}{c|}{\textbf{2/4 overlap anchor}} & \multicolumn{2}{c|}{\textbf{3/4 PASS}} & \multicolumn{2}{c|}{\textbf{0/4 PASS}} & 0/1 measured \\
    \end{tabular}
    \caption{\textbf{Main structure-gate readout: 4 Sherwood feedback variants ($P_1$--$P_4$) $\times$ 3 primary gates.} Configuration: \texttt{pub-t1}, $50{,}000$ steps, $n_{\text{rays}}=64$, seed $=0$, $\langle F\rangle_{\text{obs}}=\AnchorMeanF$~\cite{kirkman2007}. KS and $\langle F_{\rm pred}\rangle$ are reported with 5-seed sightline-level bootstrap ($K=\BootstrapK$ resamples, prediction and truth-side seeds from independent PCG64 streams; anti-degeneracy F1/F2 audit PASS, suppl.); $P_F$ residual reported at the seed=42 single-seed anchor. \textbf{Verdict: mean-flux 2/4 CIs overlap the Kirkman+2007 anchor at q84 ($P_3$/$P_4$ marginal); KS 3/4 PASS bootstrap-robustly ($P_2$ q84 $=\KSBootstrapQHiPtwo$ fails $1.5\times$); $P_F$ 0/4 PASS, uniformly $\PFfactorOverGateLo\times$--$\PFfactorOverGateHi\times$ over the $\GatePF$ bar.} $^\dagger$The 1D-along-ray density-correlation proxy $r_\rho^{\log}$ is a \emph{necessary-but-not-sufficient surrogate} for the full 3D $\xi_{\hat\rho,\rho}(r=2\,h^{-1}\,\text{Mpc})>\GateXi$ gate; the value shown is the cost-survey T3 checkpoint at $P_1$ (anchor-invariant); \texttt{pub-t1} re-evaluation owed and sequenced behind \S\ref{sec:nextsteps}.}
    \label{tab:headline-3gate}
\end{table*}

\subsection{Cosmological evaluation: $P_F$ falsifies the rescale-preview reading}
\label{sec:experiments:cosmological-eval}
The three structure gates ($P_F(k_\parallel)$, $\xi_{\hat\rho,\rho}(r)$, KS-distance on the $F$-PDF) are evaluated at two checkpoints: (i) the Tier-3 cost-survey checkpoint at step $10{,}000$, the data available before the corrected-anchor run, and (ii) the Tier-1 corrected-anchor full-schedule \texttt{pub-t1} run at step $50{,}000$ across all four feedback variants (MLflow runs \texttt{31acdf9d\dots} / \texttt{f7fafa23\dots} / \texttt{62aeb93a\dots} / \texttt{fc3817b3\dots} for $P_1$/$P_2$/$P_3$/$P_4$). Two flux-domain gates ($P_F$, $F$-PDF) are evaluated directly on each checkpoint; the line-of-sight density axis is reported via the 1D-along-ray proxy (per-sightline Pearson $r$ on $\log_{10}\rho/\langle\rho\rangle$ at integrator quadrature points), a necessary-but-not-sufficient surrogate for the full 3D $\xi_{\hat\rho,\rho}(r)$.

\textbf{Rescale-as-preview, retracted.} The cost-survey rescale-as-preview interpretation is empirically falsified by the trained-at-target $P_F$ residual (overlay figure in suppl.): $4\times$ more training and the corrected anchor increased $|\Delta P_F/P_F|$ from $\PFrescaledPercent$ (rescaled preview) to $\PFresidPonePercent$ at $P_1$.

\textbf{Empirical observation and diagnosis: $P_F$ is structurally bound, not calibration-bound.} At the corrected anchor and full publication schedule (per-cell values in suppl.), the four-physics bundle at the seed=42 single-seed anchor achieves: $\langle F_{\text{pred}}\rangle \in [\meanFseedFortyTwoRangeLo, \meanFseedFortyTwoRangeHi]$ in $4/4$ cells (cross-physics spread $\meanFseedFortyTwoSpread$, $<\!1\%$ bar); KS$\,< \GateKS$ in $3/4$ cells (range $[\KSseedFortyTwoPone, \KSseedFortyTwoPtwo]$, $P_2$ the single breach); pre-registered safety rails $4/4$; \texttt{peak\_vram}$=\peakVRAMPubtone$ GB; \texttt{tau\_amp}$\in[\tauampPubtoneRangeLo, \tauampPubtoneRangeHi]$; and $|\Delta P_F/P_F| \in [\PFresidLo, \PFresidHi]$ over $k_\parallel \in [10^{\PFkLoExp}, 10^{\PFkHiExp}]$ s/km in $4/4$ cells, $\PFfactorOverGateLo\times$--$\PFfactorOverGateHi\times$ over the $\GatePFdec$ gate. The 5-seed sightline-level bootstrap returns $\langle F_{\rm pred}\rangle \in [\meanFBootstrapRangeLo, \meanFBootstrapRangeHi]$ (cross-physics spread $\meanFBootstrapSpread$, systematically shifted below the seed=42 anchor by $\meanFseedDeviationRangeLo$--$\meanFseedDeviationRangeHi\sigma$); KS closes $3/4$ bootstrap-robustly. The conjunction --- mean-flux closure at the single-seed anchor, near-uniform KS closure, and a uniformly $\sim\!4\times$-over $P_F$ residual --- is inconsistent with a calibration deficit (which would shift mean-flux without preserving the $\meanFBootstrapSpread$ cross-physics tightness) and with a $P_2$-only seed-collapse story (which would not produce $P_F$ residuals at the three non-collapsed cells). The binding gap is in second-order spectroscopic statistics that survive matched mean-flux and matched flux PDF.

\textbf{Decoupling finding.} Log-space $\tau$-MSE is not a sufficient proxy for the structure-gate spectroscopic statistics; see suppl.

Companion plots and per-cell tables are in the supplementary.

\subsection{$\tau$-binned residual decomposition: saturation-regime under-fitting, positively identified cross-physics}
\label{sec:experiments:tau-binned}
The structural-$P_F$ verdict of \S\ref{sec:experiments:cosmological-eval} states \emph{that} the gap is in second-order spectroscopic statistics and \emph{not} in calibration. The diagnostic question is \emph{where on the $F$ axis} the gap lives. We answer this with a $\tau$-binned residual decomposition (\texttt{scripts/tau\_binned\_residual.py --cell all}) over the four \texttt{pub-t1} corrected-anchor checkpoints at step $50{,}000$, seed $=42$, $1024$ rays. The diagnostic partitions the rendered $F$ axis into seven bins matching the saturated-absorber definition of \S\ref{sec:method:loss} --- linear regime $F > 0.9$ (bins 1--4, $\sim\!99.6\%$ of pixels), transition band, and saturation band $F \in [\SatBandLo, \SatBandHi]$ (bin 7) --- and reports the per-bin residual mass $\sum |\Delta F|$ alongside pixel share. The concentration ratio $R = (\text{residual-mass share}) / (\text{pixel-mass share})$ measures how much of the per-sightline error budget the saturation band absorbs relative to its size.

\textbf{Finding.} The saturation-band $R$ is $\{\RsatPone, \RsatPtwo, \RsatPthree, \RsatPfour\}$ across $P_1$/$P_2$/$P_3$/$P_4$ (per-cell breakdown and 4-panel residual-mass histograms in suppl.). All four cells exceed the positive-ID floor $R \geq \GateRsatFloor$ by $> \!\RsatFactorOverFloor\times$, the cross-physics spread is $\sim\!\RsatSpreadRelative$ relative, and $P_4$ (strongest-feedback variant, where cool dense structure is most prevalent) has the largest $R$. The linear regime contributes $|\langle \Delta F\rangle| \leq \RsatLinearBandResidual$ across all four cells, so the cross-physics flux residual budget is dominated by the saturation band rather than spread across the $F$ axis. Within the saturation band, the per-bin Pearson $r$ on $\tau$ between predicted and ground-truth values is $r_{\text{pearson}}(\tau) = \RsatPearsonTauPone$ at $P_1$: the $\tau$-rank order collapses where it matters most for $P_F$, ruling out a global $\tau$ rescaling (uniform $\tau \to s \cdot \tau$ preserves rank order in-band) as a sufficient fix.

\textbf{Interpretation under the falsification discipline.} The decomposition positively identifies the saturation regime as the residually-supported mechanism cross-physics. We do not claim closure of the inferential gap: the strict ``saturation under-fitting \emph{is the cause} of the $P_F$ gap'' claim is reserved for the successor counterfactual run (\S\ref{sec:nextsteps:pf}). The current claim ceiling is \emph{positively identified as the residually-supported mechanism cross-physics; counterfactual closure (saturation-aware loss) deferred to successor work}.

The full-resolution per-physics breakdown (4-panel residual-mass histograms with saturation-band shading) is provided as Fig.~3 of the supplementary. The per-cell $R$ values are tabulated there alongside the figure.

\subsection{Substrate-discriminability probe at the reconstruction scale}
\label{sec:experiments:substrate-probe}
The structure-gate readout (\S\ref{sec:experiments:headline}) and the $\tau$-binned residual decomposition (\S\ref{sec:experiments:tau-binned}) characterize the rendered $F$ axis. A separate question is what cross-physics signal the truth density field carries at the spatial scale a downstream 3D consumer would slice it on. We answer this with a probe-classifier construction~\cite{alain_bengio2017probing}: a 3D ResNet-18 (halved channels, $\SubstrateProbeParams$ parameters) is trained to assign $\SubstrateProbeCropSize^3$-voxel crops of the truth $\rho/\bar\rho$ field --- $\SubstrateProbeVoxelKpc$ kpc/h voxels, a $\SubstrateProbeBoxMpc\,h^{-1}$ Mpc cube edge, matched to the $\xi_{\hat\rho,\rho}$ evaluation scale of \S\ref{sec:experiments:headline} --- to one of the four Sherwood feedback variants $\{P_1, P_2, P_3, P_4\}$. The classifier accuracy is a lower bound on the cross-physics discriminability the substrate carries; the construction is read as a measurement of representational signal at this scale, not as a paper claim about Sherwood-recipe distinguishability per se (which is published in 1D flux statistics; see below).

The probe is paired with a two-scalar $[\text{mean}, \text{variance}]$ MLP baseline as a pre-registered anti-degeneracy floor: any signal the 3D ResNet exploits beyond the marginal one-point density distribution must clear a $\SubstrateProbeADfiveBarPP$ pp margin over this baseline (Geirhos et al.~\cite{geirhos2020shortcut} shortcut-learning audit). Confidence intervals on $\hat A$ are reported in two forms --- ordinary $K{=}1{,}000$ resample bootstrap and a block bootstrap~\cite{politis_romano1994} at block size $\SubstrateProbeBlockVoxels$ voxels ($\sim\!2\times$ the gate scale) to respect the spatial autocorrelation of the underlying field. Evaluation is on $N_{\text{crops}} = \SubstrateProbeNcrops$ held-out crops drawn from an axis-0 70/15/15 spatial split.

\textbf{Result.} The 3D ResNet reaches $\hat A_{\text{overall}} = \SubstrateProbeAcc$ on the 4-way task; the two-scalar baseline reaches $\SubstrateProbeMeanVarAcc$, against a chance baseline of $\SubstrateProbeChance$. The margin is $\SubstrateProbeMarginPP$ pp, below the pre-registered $\SubstrateProbeADfiveBarPP$ pp anti-degeneracy floor, and we therefore decline to publish $\hat A_{\text{overall}}$ as a primary accuracy number for the substrate. Block-bootstrap and ordinary CIs on $\hat A_{\text{overall}}$ are $[\SubstrateProbeCIblockLo, \SubstrateProbeCIblockHi]$ and $[\SubstrateProbeCIordLo, \SubstrateProbeCIordHi]$ respectively (maximum CI-lower-bound difference $0.004$, $\ll 0.02$; the block correction does not change the conclusion). Per-quintile accuracies on the val-set-fixed equal-occupancy distance bins are $\hat A(r_{25}) = \SubstrateProbeAccQone$, $\hat A(r_{50}) = \SubstrateProbeAccQtwo$, $\hat A(r_{75}) = \SubstrateProbeAccQthree$ --- a flat curve with cross-quintile range $\sim\!\SubstrateProbeAccRangePP$ pp. The confusion matrix shows the classifier doing a $P_1$-vs-$P_4$ binary discrimination at reasonable confidence and chance-level guessing on the intermediate $P_2$/$P_3$ variants: the extremes (no feedback vs.\ strong AGN) are separable at this substrate; the middle (stellar wind vs.\ wind$+$AGN) is not.

\textbf{Prior-art refinement.} Cross-physics discriminability of the Sherwood $\{P_1$--$P_4\}$ variants is established at the 1D Ly$\alpha$-forest flux-statistic scale by the suite-overview comparisons in Bolton et al.~\cite{bolton2017sherwood} and by the cross-physics $P_F$ differentials reported in Iršič et al.~\cite{irsic2017xq100}. The present result narrows the picture: at the $\SubstrateProbeCropSize^3$ density-crop substrate matched to the gate scale, the discriminative signal collapses to what two summary moments already capture. The implication is not that the variants are physically indistinguishable but that the 3D $\rho$ substrate at this resolution is not the substrate on which higher-order signal lives. A successor probe at a different substrate scale or on a different physical field is the natural follow-on (\S\ref{sec:nextsteps}).

\subsection{Open evaluation axes and compute footprint}
\label{sec:experiments:compute}
Two evaluation axes remain open against the \texttt{pub-t1} checkpoints. (a) The 1D-along-ray density correlation $r_\rho^{\log}$ has not been re-evaluated on the corrected-anchor checkpoints (the cost-survey value $\rrhoLogProxyPone$ at $P_1$ is anchor-invariant by construction; the \texttt{pub-t1} cross-physics row is owed and sequenced behind the saturation-band diagnostic). (b) The full 3D $\xi_{\hat\rho,\rho}(r=2\,h^{-1}\,\text{Mpc}) > \GateXi$ requires the \texttt{SherwoodIGM\_gal} HDF5 particle snapshots ($\sim 40$ GB) for chunked CIC ground-truth reconstruction --- the dataset has landed on Juno scratch and the evaluator is ready, but the gate awaits a Tier-4 checkpoint not yet in hand. Validation, Tiers 1--3 across four variants, and the four-cell \texttt{pub-t1} bundle have consumed approximately \GPUHoursSpent\ GPU-hours; the remaining four Tier-4 cells require an estimated $\sim\!\GPUHoursTierFour$ GPU-hours, dominated by the $n_{\text{rays}}=16{,}384$ corner and currently blocked on the saturation-band mechanism call (\S\ref{sec:nextsteps:tier4}).
