\section{Discussion and Close-Out}
\label{sec:nextsteps}

This characterization closes here. The explicit-grid probe (\S\ref{sec:experiments:gridprobe}) showed the $z = 0.3$ flux inverse problem is under-constrained under this forward model: a maximal-capacity field fits the observed flux about $\KTwoMarginFactor\times$ better than the true field does, while recovering only weak 3D structure. Escaping the optimization collapse is necessary but not sufficient. A scaling build-out (higher resolution, multi-physics) is \emph{not} warranted by this result --- the bottleneck is the information content of the $z = 0.3$ forest, not representational capacity or compute --- so we do not open it. The full set of $\NumInterventions$ retired interventions, across four axes, is summarized in Tab.~\ref{tab:interventions}; the four on the saturation-band axis are detailed below as the constraints they leave behind, not as an open ranking.

\begin{table}[t]
    \centering
    \footnotesize
    \setlength{\tabcolsep}{4pt}
    \renewcommand{\arraystretch}{1.0}
    \begin{tabular}{@{}p{0.30\linewidth}p{0.20\linewidth}p{0.42\linewidth}@{}}
        \toprule
        \textbf{Intervention} & \textbf{Axis} & \textbf{Outcome} \\
        \midrule
        Saturation-aware $P_F$ band loss & loss (integrated) & FAIL --- constant-prediction collapse \\
        FGPA-tail per-pixel regularizer & loss (per-pixel) & FAIL --- constant-prediction collapse \\
        Velocity-gradient conditioning & architecture input & FAIL at smoke --- density-head collapse \\
        Physics-id embedding (joint) & data (pooled) & FAIL at smoke --- trivial-flux collapse \\
        $L_1/L_2$ $P_F$-MSE levers ($\times 5$: lr, per-task clip, reduction, $k$-norm) & loss & all retire to the variance-collapse basin \\
        fGPA-residual density prior & architecture / target & falsified ($120\times$ under the feasibility boundary) \\
        Direct $\rho$-MSE pretraining & supervision target & falsified; active variance shrinkage (1-of-3 lr cells) \\
        Skip-rich MLP body & architecture & falsified; $n{=}10$ seeds, all negative \\
        Linear $\log\rho$ head (no Softplus) & parameterization & collapse at probe scope \\
        \bottomrule
    \end{tabular}
    \caption{\textbf{The $\NumInterventions$ retired interventions across four axes (loss, architecture, data, parameterization), $z{=}0.3$.} Each was specified before its run and retired at a pre-committed stop condition; none closed the $P_F$ gate. Two further checks (microbatch size, physics variant) are diagnostics, not interventions. \emph{Verdict:} the $P_F$ deficit is not resolved by added expressiveness on any tested axis --- the bottleneck is the information content of the low-$z$ forest, not the neural representation.}
    \label{tab:interventions}
\end{table}

\subsection{Four retired counterfactual interventions on the saturation-band axis}
\label{sec:nextsteps:pf}
Each entry below was specified before its run, executed under a pre-committed stop condition, and retired at the indicated tier when the empirical signature did not match the designed minimum. The four failures share a structural cause: any regularizer evaluated only on the network's self-consistent state is gameable when the data loss is weakly informative on the diffuse-bin majority, because constant-field collapse trivially satisfies any algebraic relation among the predicted fields --- a shortcut-learning mode in the sense of Geirhos et al.~\cite{geirhos2020shortcut}.

\textbf{1.\ Saturation-aware $P_F$ band loss --- retired.} An integrated relative-residual band loss $((P_{F,\text{pred}}-P_{F,\text{truth}})/P_{F,\text{truth}})^2$ was tested on $P_1$ at the cost-survey schedule ($n_{\text{rays}}=64$, $12{,}500$ steps, seed $=0$, anchor $\AnchorMeanF$). Eval $|\Delta P_F/P_F|=\PFdfortyEval$ ($+\PFdfortyEvalDelta$ worse than the \texttt{pub-t1} $P_1$ baseline of $\PFresidPone$); KS gate broke ($\KSdfortyEval$). Degeneracy class: amplitude-shrink with shape preservation ($P_{F,\text{pred}}\!\approx\!0.43\,P_{F,\text{truth}}$ uniformly), a constant-prediction minimum of the integrated relative-residual formulation when $P_{F,\text{truth}}$ is small in any band.

\textbf{2.\ FGPA-tail per-pixel regularizer --- retired.} A per-pixel Huber penalty on the FGPA scaling $\log\tau \propto \beta\log(\rho/\bar\rho)+\gamma\log T$~\cite{bi_davidsen1997,hui_gnedin1997} (Hui--Gnedin $\beta\!=\!1.6$, $\gamma\!=\!-0.7$) was tested at smoke ($P_1$, 50 steps) and confirmed at Tier-1 ($25{,}000$ steps, same configuration). The training-time regularizer descended $20\times$ ($6.37 \to 0.31$), but eval $|\Delta P_F/P_F| = 1.000$ ($\sim\!10\times$ worse than baseline) and the FGPA-valid input fields collapsed to near-uniform constants on the Tier-1 checkpoint (density $\approx 71.5$, $X_{HI}\!\approx\!3.3\!\times\!10^{-5}$). Degeneracy class: constant-prediction collapse on a self-consistent state --- constants $(\rho, T)$ trivially satisfy the FGPA single-equation constraint, and the mean-flux anchor pulls the resulting constant $\tau$ to mean-$F\!\approx\!1$.

\textbf{3.\ Velocity-gradient conditioning branch --- retired at smoke.} A conditioning branch on the observation-derived $\partial v_{\text{pec}}/\partial \chi$ along the line of sight was injected into the density head, chosen specifically to break the self-consistent-state degeneracy of \#2 (ground-truth-anchored, not network-predicted). The 50-step $P_1$ smoke passed gates 1--4 and 6 but failed gate 5 (predicted-density spread $0.0071$ vs.\ $\geq 1.45$ bar, $200\times$ below floor); the density head collapsed to $\approx 0$ while the $X_{HI}$ head retained structure. Degeneracy class: partial-collapse signature distinct from \#1 (integrated-amplitude shrink) and \#2 (full self-consistent collapse). Retired per pre-committed stop condition without Tier-1 dispatch.

\textbf{4.\ Joint-physics conditional MLP with physics\_id embedding --- retired at smoke.} A 16-dim physics\_id embedding was concatenated into the MLP input to train one network jointly on sightlines pooled across $\{P_1, P_2, P_3, P_4\}$ (4096 sightlines, 1024 per physics, microbatch interleaving), targeting the conditioning axis (cross-physics pooling via an added conditioning input) orthogonal to the loss/regularizer/architecture-input axes of \#1--\#3. The 50-step P-mixed smoke passed gates 1, 4, 6, 7 but failed gates 2, 3, 5: loss-flat (ratio $1.0036$ vs.\ $<0.85$ required), per-physics mean-$F\!=\!1.000$ in all four cells (trivial-flux collapse), and per-physics predicted-density spread $6.4\!\times\!10^{-5}$ to $2.0\!\times\!10^{-4}$ ($\sim\!10{,}000\times$ below the $1.45$ floor). The embedding itself was non-degenerate (gate 7 PASS, max pairwise $L_2 = 7.045$ in 16-dim space); the network learned 4 distinct codes yet routed all of them into the same trivial-flux density-collapse solution. Degeneracy class: combined-trivial-collapse-with-active-embedding, distinct from \#1 (amplitude-shrink), \#2 (self-consistent constant collapse), and \#3 (asymmetric-head partial collapse), and notable for reproducing the constant-field-collapse mechanism of \#1/\#2 even when the conditioning surface is non-degenerate.

The four interventions on loss, regularizer, architecture-input, and conditioning axes cataloged four distinct degeneracy signatures, all reachable from the log1p$+$cap$+$mask supervision regime with the Kirkman~\cite{kirkman2007} $\langle F\rangle$ anchor. The reading is underspecification of that supervision target~\cite{damour2020underspecification}, which the explicit-grid probe then confirmed at the inverse-problem level: even a maximal-capacity field does not recover the true structure from this flux supervision.

\subsection{Scope and what we do not open}
\label{sec:nextsteps:scope}
The result is scoped to the Sherwood Suite at $z = 0.3$, $P_1$. We do not claim neural IGM tomography is impossible: at $z \approx 2$--$3$ the forest is information-rich and CLAMATO~\cite{lee2018clamato} and TARDIS~\cite{horowitz2019tardis} reconstruct 3D density from sparse skewers. The claim is narrower --- at $z = 0.3$, this method class hits a characterized under-constraint wall, and the wall's location is explained by the information content of the low-$z$ forest, not by a deficiency of the architecture. A scaling build-out (higher resolution, multi-physics) is not warranted, because the explicit-grid probe shows the bottleneck is information content, not capacity. We considered adopting a different autodiff engine (JAX) for such a build-out; with no build-out opened, the migration is not adopted ([D-74]). A scoped diagnostic spike --- porting the forward model plus one Hessian/Fisher diagnostic to harden the under-determination claim --- remains an option if elected, but is not committed and not in flight. The corrected per-lag $\xi$ coefficient $\xi_{\text{cross}}/\sqrt{\xi_{\hat\rho\hat\rho}\,\xi_{\rho\rho}}$ is banked as future work: it needs the grid cube and remains frame-mismatch-confounded (\S\ref{sec:experiments:gridprobe}). The substrate-discriminability probe (\S\ref{sec:experiments:substrate-probe}), which fell below its anti-degeneracy floor at the $\SubstrateProbeCropSize^3$ scale, likewise leaves substrate scale, alternative physical fields, and reconstructed-substrate evaluation as the load-bearing axes for any successor probe (suppl.).
