\section{Related Work} % Project Foundation: Background for Stages 1-3
\label{sec:related}

\textbf{Cosmological Ly$\alpha$ tomography.} The reconstruction of the 3D IGM from sparse Ly$\alpha$ forest sightlines has a $\sim\!25$-year methodological history. Pichon et al.~\cite{pichon2001tomography} cast the inversion as a nearest-neighbor / Wiener-style filter on quasar sightlines, establishing the voxel-grid framing that the modern era inherited. Stark et al.~\cite{stark2015tomography} formalized the sparse-tomography setting at $z \sim 2.3$ and demonstrated Wiener-filter reconstruction at survey-realistic inter-sightline spacing; Horowitz et al.~\cite{horowitz2019tardis} introduced TARDIS, a constrained-optimization scheme that absorbs absorption-line physics directly into the inverse-problem objective. Lee et al.~\cite{lee2015, lee2018clamato} consolidated the observational requirements and first-light realization of the CLAMATO survey at $z\sim 2.3$. These are the comparison points for any continuous-representation work at low redshift. They share a common design choice --- an explicit voxel grid as the reconstruction primitive --- which the present work substitutes with a continuous coordinate-conditioned MLP.

\textbf{Implicit neural representations.} In computer vision, NeRF~\cite{mildenhall2020nerf} initiated a wave of coordinate-based MLP scene representations supervised through a differentiable rendering integral. The bandwidth of the input encoding turns out to control representable spatial frequency in a precise way that Tancik et al.~\cite{tancik2020fourier} characterize for Fourier features; SIREN~\cite{sitzmann2020siren} extends the same idea with periodic activations throughout the MLP, with sharper-than-ReLU spectral profiles on high-frequency targets. The representation has been ported beyond static scenes: NeRV~\cite{chen2021nerv} represents whole videos as a single coordinate-conditioned MLP, and Instant NGP~\cite{muller2022instant} uses a multiresolution hash grid to accelerate the same family of architectures by orders of magnitude. We adopt Fourier positional encoding at $L=10$ as our bandwidth selector (Sec.~\ref{sec:method}); the higher-bandwidth-without-grid alternatives of~\cite{sitzmann2020siren, muller2022instant} are open ablation axes for the successor design.

\textbf{Physics-informed neural networks.} The introduction of physical constraints into neural-network optimization --- typically as soft penalty terms enforcing PDE residuals --- is the PINN framework of Raissi et al.~\cite{raissi2019pinn}, surveyed in Karniadakis et al.~\cite{karniadakis2021pinns}. Our setup is structurally related but operationally distinct: rather than penalizing residuals of a governing equation, we propagate gradients through a complete physical forward model (the RSD-convolved Voigt-Hjerting kernel, Sec.~\ref{sec:method}) and supervise against an observable that is the model's natural output. The FGPA $\tau \propto \rho^{1.6} T^{-0.7}$~\cite{bi_davidsen1997, hui_gnedin1997} is the analogous PDE-residual prior in our domain; it figures in the successor-work design (Sec.~\ref{sec:nextsteps:pf} \#1) but is not a load-bearing component of the present model. Earlier cosmological ML work used CNN and GAN architectures over voxelized inputs~\cite{zamudio2019higan, ravanbakhsh2017cosmology} without coupling to a differentiable observational operator.

\textbf{Sparse-view inverse rendering in medical imaging.} The closest cross-domain analogue of our problem is sparse-view cone-beam computed tomography (CBCT) reconstructed via implicit neural fields. MedNeRF~\cite{corona2022mednerf} adapts NeRF for single-X-ray 3D-aware reconstructions; NAF~\cite{zha2022naf} replaces voxel grids with a Neural Attenuation Field for sparse-view CBCT and demonstrates the same memory-vs-frequency tradeoff our setting confronts; TensoRF~\cite{chen2022tensorf} explores a hybrid factorized representation that interpolates between grid and INR. Two structural differences distinguish our setting from medical sparse-view tomography: (i) the forward operator is wavelength-resolved spectroscopic absorption with explicit kinematic content (peculiar-velocity Doppler shifts and thermal Doppler broadening), not a frequency-integrated attenuation integral; and (ii) the projection geometry is anisotropic pencil-beam (1D rays through a 3D volume with no angular diversity), where medical CBCT enjoys multi-angle coverage. These differences are why a straight port of NAF or MedNeRF is not a baseline for this paper, and why \cite{stark2015tomography} and \cite{horowitz2019tardis} are the load-bearing comparisons.

\textbf{Low-redshift Ly$\alpha$ mean-flux anchor.} The mean-flux anchor at $z = 0.3$ is sourced from Kirkman et al.~\cite{kirkman2007}, ``Continuous statistics of the Ly$\alpha$ forest at $0 < z < 1.6$'', whose Table 2 and Section 4 power-law fit $D_A(z) = 0.016(1+z)^{1.01}$ yield $\langle F\rangle(z=0.3) = 0.979 \pm 0.005$. The HST/COS low-$z$ survey of Danforth et al.~\cite{danforth2016} (82 UV-bright AGN sightlines, 2611 H I absorption systems at $z < 0.85$) is retained as the canonical column-density-distribution reference but does not directly tabulate $\langle F\rangle(z)$. Faucher-Gigu\`ere et al.~\cite{faucher2008meanflux} measure $\tau_{\text{eff}}(z)$ over $2 \leq z \leq 4.2$ from high-resolution quasar spectra and is retained as a precedent for the mean-flux estimator at higher redshift. The Phase~A.1 audit (LEDGER [D-34], 2026-05-08) corrected the anchor value: an earlier draft attributed $0.877$ to the $z=0.3$ regime; that is the FG+2008 high-$z$ value, mislabeled in project history.

\textbf{DLA and Lyman-limit-system exclusion.} Damped Lyman-alpha systems (DLAs, $N_{HI} \geq 2 \times 10^{20}$ cm$^{-2}$~\cite{wolfe2005}) and strong Lyman-limit systems (LLS, $N_{HI} \gtrsim 10^{17.2}$ cm$^{-2}$~\cite{prochaska2010}) are conventionally masked out of forest analyses because $F = e^{-\tau}$ saturates and carries no recoverable structure. Our $\tau > 10^5$ saturated-core threshold corresponds to $N_{HI} > 1.9 \times 10^{18}$ cm$^{-2}$ at $T = 10^4$ K, an order of magnitude below the DLA boundary and above the strong-LLS regime, so the mask catches both classes by a wide margin.
